%\begin{doublespace}

\chapter*{Preface}

%\vspace{-50mm}

%{\Large Preface} \\

This book provides an introduction to statistics and its applications in the life sciences and biomedical research.  It is based on the freely available \textsl{OpenIntro Statistics, Third Edition}, and, like \textsl{OpenIntro} it may be downloaded as a free PDF at \url{https://github.com/OI-Biostat/oi_biostat_text}.  The text adds substantial new material, eliminates sections from \textsl{OpenIntro}, and re-uses some material directly. We have retained some of the examples and exercises from \textsl{OpenIntro} that may not come directly from medicine or the life sciences but illustrate important ideas or methods. 

\textsl{Introduction to Statistics for the Life and Biomedical Sciences} is intended for undergraduate and graduate students interested in careers in biology or medicine, and may also be profitably read by students of public health.  It covers many of the traditional introductory topics in statistics used in those fields, in addition to discussing some newer methods being used in molecular biology. Statistics has become an integral part of research in medicine and biology, and the tools for displaying, summarizing and drawing inferences from data are essential both for understanding the outcomes of studies and for incorporating measures of uncertainty into that understanding.  An introductory text in statistics for students considering careers in medicine, public health, or the life sciences should be more than simply the usual introduction, supplemented with the occasional example from biology or medical science. We have worked to develop a text centered around published data from biology and medicine that conveys the value of statistics in medical and biological research, in addition to covering topics that are of particular importance to students in biology, medicine, and public health.

\begin{comment}
Along with the value of careful, robust analyses of experimental and observational data, it should convey some of the excitement of discovery that emerges from the interplay of science with data collection and analysis. We hope we have conveyed some of that excitement here.
\end{comment} 

We have tried to balance the sometimes competing demands of mastering the technical aspects of analysis methods with gaining an understanding of important concepts.  The examples and exercises include opportunities for students to build skills in conducting data analyses and to state conclusions with clear, direct language that is specific to the context of a problem.  We also believe that computing is an essential part of statistics, just as mathematics was when computing was more difficult or expensive.  The text includes many examples where software is used to aid in the understanding of the features of a dataset as well as exercises where computing is used to help illustrate the notions of randomness and variability. We use the freely available \textsf{R} statistical language with the \textsl{R Studio} interface.  Information on downloading \textsf{R} and \textsl{R Studio} may be found at \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}.  Nearly all examples and exercises can be adapted to either SAS, Stata or other software, but we have not done that.

The datasets used in this book are available via the \textsf{R} \texttt{openintro} package available on CRAN\footnote{Diez DM, Barr CD, \c{C}etinkaya-Rundel M. 2012. \texttt{openintro}: OpenIntro data sets and supplement functions. \urlwofont{http://cran.r-project.org/web/packages/openintro}.}  and the \textsf{R} \texttt{oibiostat} package available via \href{<https://github.com/OI-Biostat/oi_biostat_data>}{GitHub}.

%\textit{DH: This is where we should mention the parallel book of examples in R, reproducing the figures and tables in the book (without adornment) and providing a bit of instruction in R.}



\begin{comment}

This book may be downloaded as a free PDF at \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}. %This download is free and need not be supplemented by the purchase of a paperback copy.
\vspace{3mm}

\noindent We hope readers will take away three ideas from this book in addition to forming a foundation of statistical thinking and methods.\vspace{-1mm}
\begin{enumerate}
\setlength{\itemsep}{0mm}
\item[(1)] Statistics is an applied field with a wide range of practical applications.
\item[(2)] You don't have to be a math guru to learn from real, interesting data.
\item[(3)] Data is messy, and statistical tools are imperfect. But, when you understand the strengths and weaknesses of these tools, you can use them to learn about the real~world.
\end{enumerate}
%This textbook is supplemented by many free, online resources to help students apply the methods they learn in this textbook and beyond.

\end{comment}

\subsection*{Textbook overview}

The chapters of this book are as follows:

\begin{description}
\setlength{\itemsep}{0mm}

%\begin{itemize}
\item[1. Introduction to data.] Data structures, basic data collection principles, numerical and graphical summaries, and exploratory data analysis.
\item[2. Probability.] The basic principles of probability.
\item[3. Distributions of random variables.] Introduction to random variables and distributions of discrete and continuous distributions.
\item[4. Foundations for inference.] General ideas for statistical inference in the context of estimating a population mean.
\item[5. Inference for numerical data.] Inference for one-sample and two-sample means with the $t$ distribution, power calculations for a difference of means, and ANOVA.
\item[6. Simple linear regression.] An introduction to linear regression with a single explanatory variable, evaluating model assumptions, and inference in a regression context.
\item[7. Multiple linear regression.] General multiple regression model, categorical predictors with more than two values, interaction, and model selection.
\item[8. Inference for categorical data.] Inference for proportions using the normal and chi-square distributions, as well as simulation and randomization techniques.
%\item[9. Logistic regression.] description

%\end{itemize}
\end{description}

%\textbf{The remainder of this section requires revision}

%This textbook was written to allow flexibility in choosing and ordering course topics. The material is divided into two pieces: main text and special topics. The goal of the main text is to allow people to move towards statistical inference and modeling sooner rather than later. Special topics, labeled in the table of contents and in section titles, may be added to a course as they arise naturally in the curriculum.

\emph{OpenIntro Statistics} was written to allow flexibility in choosing and ordering course topics. The material is divided into two pieces: main text and special topics. The main text has been structured to bring statistical inference and modeling closer to the front of a course. Special topics, labeled in the table of contents and in section titles, may be added to a course as they arise naturally in the curriculum.

%\emph{OpenIntro Statistics} would also serve as a helpful supplement in a course preparing students for the Advanced Placement Statistics exam, either though the textbook or use of the online resources outlined below. %We hope for this textbook to be deemed ``AP ready'' for future editions, which will require an application for approval from College Board. %The AP material would include the main and optional materials of Chapters~1-7 with the possible omission of Sections~1.8, 3.5, 4.7, 6.3-6.4, and~7.4.

\subsection*{Examples, exercises, and appendices}

Examples and within-chapter exercises throughout the textbook may be identified by their distinctive bullets:

\begin{example}{Large filled bullets signal the start of an example.}
Full solutions to examples are provided within the main text and often include an accompanying table or figure.
 \end{example}

\begin{exercise}
Empty bullets signal to readers that an exercise has been inserted into the text for additional practice and guidance. Solutions are provided for all within-chapter exercises in footnotes.\footnote{Full solutions are located in the footnotes.}
\end{exercise}

There are exercises at the end of each chapter that are useful for practice or homework assignments. Solutions are in Appendix~\ref{eoceSolutions}. %These end-of-chapter exercises are also available online in a public question bank at \textbf{openintro.org}, and the available selection is constantly growing based on teacher contributions. Numbered citations in end-of-chapter exercises may be found in Appendix~B.

Probability tables for the normal, $t$, and chi-square distributions are in Appendix~\ref{distributionTables}, and PDF copies of these tables are also available from \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}} for anyone to download, print, share, or modify.

\subsection*{OpenIntro, online resources, and getting involved}

OpenIntro is an organization focused on developing free and affordable education materials. \emph{OpenIntro Statistics}, our first project, is intended for introductory statistics courses at the high school through university levels.

%We encourage anyone learning or teaching statistics to visit \textbf{openintro.org} and get involved by using the many online resources, which are all free and rapidly expanding, or by creating new material. Students can test their knowledge with practice quizzes for each chapter, or try an application of concepts learned in each chapter using real data and the top-rated and free statistical software \emph{R}. Teachers can download the source for course materials, labs, slides, data sets, R figures, or create their own custom quizzes and problem sets for students to take on the website. Everyone is also welcome to download this textbook as a PDF or the book's source files to create a custom version of this textbook or to simply share a copy with a friend or on a website. All of these products are free, and we want to be clear that anyone is welcome to use these online tools and resources with or without this textbook as a companion.

We encourage anyone learning or teaching statistics to visit \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}} and get involved. We also provide many free online resources, including free course software. 
%, or by creating new material. Students can test their knowledge with practice quizzes for each chapter, or try an application of concepts learned using real data. 
Datasets for this textbook are available on the website and through a companion \textsf{R} package.\footnote{Diez DM, Barr CD, \c{C}etinkaya-Rundel M. 2012. \texttt{openintro}: OpenIntro data sets and supplement functions. \urlwofont{http://cran.r-project.org/web/packages/openintro}.} All of these resources are free, and we want to be clear that anyone is welcome to use these online tools and resources with or without this textbook as a companion.

%Teachers can download the source files for this book, labs, data sets, or create their own custom quizzes and problem sets for students to take at \href{http://www.openintro.org}{\textbf{openintro.org}}. 
%Anyone can download a PDF or the source files of this textbook for modifying and sharing at \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}.

We value your feedback. If there is a particular component of the project you especially like or think needs improvement, we want to hear from you. You may find our contact information on the title page of this book or on the \href{http://www.openintro.org/about.php}{About} section of \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}.

\subsection*{Acknowledgements}

This project would not be possible without the dedication and volunteer hours of all those involved. No one has received any monetary compensation from this project, and we hope you will join us in extending a \emph{thank you} to all those volunteers below.

The authors of \textsl{OpenIntro Statistics} would like to thank Andrew Bray, Meenal Patel, Yongtao Guan, Filipp Brunshteyn, Rob Gould, and Chris Pope for their involvement and contributions. %A special thank you to Andrew Bray, who has developed the R labs and diligently worked to translate end-of-chapter exercises to the project website. We also deeply appreciate the contribution of Meenal Patel, who has helped raise the professional profile of OpenIntro by designing a business system and website for the project. 
We are also very grateful to Dalene Stangl, Dave Harrington, Jan de Leeuw, Kevin Rader, and Philippe Rigollet for providing us with valuable feedback.

%\textit{DH: We need to revise acknowledgements.}

% Add note about margin notes for notation

\begin{comment}

@article{smith1998fifty,
  title={Fifty years of randomised controlled trials},
  author={Smith, R},
  journal={British Medical Journal},
  volume={317},
  pages={1166},
  year={1998}
}

\end{comment}

%\end{doublespace}